---
title: "Total Hip Arthroplasty in Patients with Proximal Femoral Migration"
author: "Osama Ahmed Ibrahim"
output: word_document
---
```{r message=FALSE, warning=FALSE, include=FALSE}
knitr::opts_chunk$set(
  dpi = 300)
```

Statistical analysis:
Statistical analyses were performed using R software. Continuous variables were expressed as mean ± standard deviation or median with interquartile range, depending on distribution. Categorical variables were presented as frequencies and percentages.
Normality was assessed prior to analysis. Pre- and postoperative comparisons of paired continuous variables (limb length discrepancy and flexion deformity) were performed using the Wilcoxon signed-rank test. Repeated measures of Harris Hip Score across multiple time points were analyzed using one-way repeated-measures ANOVA, followed by post hoc pairwise comparisons with Bonferroni correction.
Changes in categorical variables (pelvic obliquity) were assessed using McNemar’s test. A p-value < 0.05 was considered statistically significant.


```{r message=FALSE, warning=FALSE, include=FALSE}

library(tidyverse)
library(readxl)
library(writexl)
library(gtsummary)
library(broom)
library(rstatix)
library(flextable)
library(patchwork)
library(glue)
library(ggsignif)
library(ggdist)


# Read Excel dataset containing arthroplasty outcomes
Arthroplasty_Data <- read_excel(path = "Project 4 data.xlsx")

# Rename key variables, create patient ID, and reorder columns
Arthroplasty_Data <- Arthroplasty_Data %>%
  rename(
    Age = age,
    Gender = gender,
    Side = side,
    Complications = complication
  ) %>%
  mutate(PatientID = row_number()) %>%
  relocate(PatientID, .before = Age)

# Recode gender variable into readable labels
Arthroplasty_Data$Gender <- factor(Arthroplasty_Data$Gender) %>%
  fct_recode("Male" = "1", "Female" = "2")

# Recode surgical side variable into labels
Arthroplasty_Data$Side <- factor(Arthroplasty_Data$Side) %>%
  fct_recode("Right" = "1", "Left" = "2")

# Standardize column names: replace "pre" with "Preoperative"
Arthroplasty_Data <- Arthroplasty_Data %>%
  rename_with(~ gsub("pre", "Preoperative", .x))

# Rename specific outcome variables for consistency
Arthroplasty_Data <- Arthroplasty_Data %>%
  rename(
    "Pelvic_Obliquity Postoperative" = `Pelvic obliquity  post`,
    "Pelvic_Obliquity Preoperative" = `Pelvic obliquity Preoperative`,
    "LLD Postoperative" = `LLD post`,
    "FD Postoperative" = `FD post`
  )

```

### Table 1. Baseline demographic and clinical characteristics of patients undergoing total hip arthroplasty 
```{r echo=FALSE, message=FALSE, warning=FALSE}
# Summary table for demographic and clinical variables
Arthroplasty_Data %>%
  select(Age, Gender, Side, Complications, `Acetabular reconstruction`) %>%
  tbl_summary(
    statistic = Age ~ "{mean} ± ({sd})"
  ) %>%
  bold_labels() %>%
  as_flex_table()
```


Results:
Baseline characteristics
A total of 24 patients undergoing total hip arthroplasty for proximal femoral migration were included in the study. The mean age was 42 ± 13 years, indicating a relatively young cohort. Males represented 63% of the study population, while females accounted for 37%. The right hip was more frequently affected (54%) compared to the left (46%). Postoperative complications were uncommon, occurring in 4.2% of patients, while 8.3% required acetabular reconstruction due to bone loss or migration.


### Table 2. Comparison of limb length discrepancy and flexion deformity before and after total hip arthroplasty 
```{r echo=FALSE, message=FALSE, warning=FALSE}
# Reshape data from wide to long format for repeated measures analysis
Arthroplasty_Data %>%
  select(starts_with("LLD"), starts_with("FD"), PatientID) %>%
  pivot_longer(
    cols = -PatientID,
    names_to = c("Measure", "Time"),
    names_sep = " ",
    values_to = "Value"
  ) %>%
  mutate(Time = factor(Time, levels = c("Preoperative", "Postoperative"))) %>%
  pivot_wider(names_from = Measure, values_from = Value) %>%
  
  # Summary table by time point
  tbl_summary(
    by = Time,
    include = c(LLD, FD),
    label = list(
      LLD ~ "Limb Length Discrepancy",
      FD ~ "Flexion Deformity"
    )
  ) %>%
  
  # Paired non-parametric test for repeated measures
  add_p(
    test = list(all_continuous() ~ "paired.wilcox.test"),
    group = PatientID
  ) %>%
  add_ci() %>%
  bold_labels() %>%
  as_flex_table()

```

Limb length discrepancy and flexion deformity
There was a statistically significant improvement in both limb length discrepancy and flexion deformity following surgery. Limb length discrepancy decreased from 2.75 (2.40–3.25) preoperatively to 0.45 (-0.30–0.70) postoperatively (p < 0.001). Similarly, flexion deformity improved significantly from 25° (15–30°) preoperatively to 5° (0–5°) postoperatively (p < 0.001), demonstrating effective correction of anatomical deformities after THA.

### Table 3. Longitudinal changes in Harris Hip Score following total hip arthroplasty 
```{r echo=FALSE, message=FALSE, warning=FALSE}
# Rename time points for clarity
Arthroplasty_Data_for_long <- Arthroplasty_Data %>%
  rename(
    "Preoperative" = `HHS Preoperative`,
    "Post 6 weeks" = `HHS post 1 (6w)`,
    "Post 6 months" = `HHS post 2 (6m)`,
    "Post 1 year" = `HHS post 3 (1y)`
  )

# Convert dataset to long format for repeated measures analysis
data_long <- Arthroplasty_Data_for_long %>%
  select(PatientID, Preoperative, `Post 6 weeks`,
         `Post 6 months`, `Post 1 year`) %>%
  pivot_longer(
    cols = c(Preoperative, `Post 6 weeks`,
             `Post 6 months`, `Post 1 year`),
    names_to = "TimePoint",
    values_to = "HHS_Score"
  ) %>%
  mutate(
    TimePoint = factor(TimePoint, levels = c(
      "Preoperative", "Post 6 weeks",
      "Post 6 months", "Post 1 year"
    ))
  )

# Perform repeated measures ANOVA
repeaetd_measures_anova <- data_long %>%
  anova_test(
    dv = HHS_Score,
    wid = PatientID,
    within = TimePoint
  )

# Extract ANOVA table
table_repeated_anova <- get_anova_table(repeaetd_measures_anova)

# Summary of HHS scores over time
data_long %>%
  tbl_summary(
    by = TimePoint,
    include = HHS_Score,
    statistic = list(all_continuous() ~ "{mean} ± {sd}"),
    label = list(HHS_Score ~ "HHS Score")
  ) %>%
  modify_header(label ~ "**Outcome**") %>%
  modify_table_body(~ .x %>%
                      mutate(p_value = style_pvalue(table_repeated_anova$p))) %>%
  modify_header(p_value ~ "**p-value**") %>%
  bold_labels() %>%
  as_flex_table()
```

Functional outcomes (Harris Hip Score)
The Harris Hip Score showed a significant and sustained improvement over time (p < 0.001). The mean score increased from 42 ± 6 preoperatively to 77 ± 3 at 6 weeks, 82 ± 3 at 6 months, and 88 ± 3 at 1 year postoperatively. The greatest improvement was observed in the early postoperative period, followed by continued but smaller gains over time, indicating progressive functional recovery.
Functional outcomes (Harris Hip Score)
Pairwise comparisons of Harris Hip Score demonstrated statistically significant differences between all time points (adjusted p < 0.001 for all comparisons). The most substantial improvement occurred between the preoperative period and 6 weeks postoperatively, while continued significant gains were observed between 6 weeks and 6 months, as well as between 6 months and 1 year.

### Table 4. Pairwise comparisons of Harris Hip Score across follow-up time points 
```{r echo=FALSE, message=FALSE, warning=FALSE}
# Pairwise comparisons between time points
posthoc_result <- data_long %>%
  pairwise_t_test(
    HHS_Score ~ TimePoint,
    paired = TRUE,
    p.adjust.method = "bonferroni"
  )

# Format post-hoc results into publication-ready table
posthoc_result %>%
  mutate(
    HHS_Comparison = paste(group1, "vs", group2),
    p.adj = gtsummary::style_pvalue(p.adj)
  ) %>%
  select(HHS_Comparison, p.adj) %>%
  flextable() %>%
  set_header_labels(
    HHS_Comparison = "Harris Hip Score Pairwise Comparison",
    p.adj = "Adjusted p-value"
  ) %>%
  theme_booktabs() %>%
  bold(part = "header") %>%
  autofit()

```


### Table 5. Changes in pelvic obliquity before and after total hip arthroplasty
```{r echo=FALSE, message=FALSE, warning=FALSE}
# Prepare paired binary outcome for McNemar test
pelvic_obl_long <- Arthroplasty_Data %>%
  select(PatientID,
         `Pelvic_Obliquity Preoperative`,
         `Pelvic_Obliquity Postoperative`) %>%
  rename(
    Preoperative = `Pelvic_Obliquity Preoperative`,
    Postoperative = `Pelvic_Obliquity Postoperative`
  ) %>%
  pivot_longer(
    cols = c(Preoperative, Postoperative),
    names_to = "Time",
    values_to = "Pelvic Obliquity"
  ) %>%
  mutate(
    `Pelvic Obliquity` = factor(
      `Pelvic Obliquity`,
      levels = c(0,1),
      labels = c("No", "Yes")
    )
  )

# Summary table + McNemar test for paired proportions
pelvic_obl_long %>%
  tbl_summary(
    by = Time,
    include = `Pelvic Obliquity`,
    type = list(`Pelvic Obliquity` ~ "categorical")
  ) %>%
  add_p(
    test = list(`Pelvic Obliquity` ~ "mcnemar.test"),
    group = PatientID
  ) %>%
  bold_labels() %>%
  as_flex_table()

```

Pelvic obliquity
Pelvic obliquity improved significantly following surgery (p = 0.008). The proportion of patients without pelvic obliquity increased from 58% preoperatively to 96% postoperatively, while those with persistent pelvic obliquity decreased from 42% to 4.2%, indicating effective coronal plane realignment after THA.


### Figure 1. Longitudinal changes in Harris Hip Score following total hip arthroplasty
```{r echo=FALSE, message=FALSE, warning=FALSE}
# Spaghetti plot with mean trend and 95% CI
ggplot(data_long, aes(x = TimePoint, y = HHS_Score, fill = TimePoint)) +
  
  # Individual patient trajectories
  geom_line(
    aes(group = PatientID),
    alpha = 0.25,
    linewidth = 0.6,
    colour = "gray40"
  ) +
  
  # Raw data points
  geom_jitter(
    alpha = 0.4,
    color = "#2C7BB6",
    size = 2,
    width = 0.2
  ) +
  
  # Mean trend line
  stat_summary(
    aes(group = 1),
    fun = mean,
    geom = "line",
    linewidth = 1.3,
    colour = "black"
  ) +
  
  # 95% confidence interval
  stat_summary(
    fun.data = mean_cl_normal,
    geom = "errorbar",
    width = 0.15,
    linewidth = 0.8,
    colour = "black"
  ) +
  
  labs(
    x = "Time Point",
    y = "Harris Hip Score"
  ) +
  
  theme_classic(base_size = 14) +
  theme(
    legend.position = "none",
    axis.text = element_text(color = "black")
  )

```
Individual patient trajectories are shown in gray. The black line represents the mean Harris Hip Score over time. Error bars indicate 95% confidence intervals.